---
title: "decoupleR Pathway Activation - VCaP DKO"
author: "Alex Watterson"
date: "2024-06-05"
output: html_document
---

#1. Load packages:
```{r}
library(tidyverse)
library(decoupleR)
library(dplyr)
library(tibble)
library(tidyr)
library(ggplot2)
library(pheatmap)
library(ggrepel)
```

#2. Reed in comparison files:
```{r}
#Data folder:
data_folder="/Users/wattersa/Downloads/decoupleR-Pathway-activation-main-main/VCaP Analysis/"

results_folder="/Users/wattersa/Downloads/decoupleR-Pathway-activation-main-main/VCaP Analysis/Results/"

#Define comparison to examine:
Comparison <- VCaP_CHD1g2_CTRL_vs_VCaP_NTC_CTRL <- read_tsv(paste(data_folder,"VCaPCHD1g2_CTRL_vs_VCaPNTC_CTRL.tsv", sep="/"), show_col_types = F)
#Comparison <- VCaP_CHD1g2_IFNg24h_vs_VCaP_NTC_IFNg24h <- read_tsv(paste(data_folder,"VCaPCHD1g2_IFNg24h_vs_VCaPNTC_IFNg24h.tsv", sep="/"), show_col_types = F)
#Comparison <- VCaP_CHD1g2MAP3K7g2_CTRL_vs_VCaP_NTC_CTRL <- read_tsv(paste(data_folder,"VCaPCHD1g2MAP3K7g2_CTRL_vs_VCaPNTC_CTRL.tsv", sep="/"), show_col_types = F)
#Comparison <- VCaP_CHD1g2MAP3K7g2_IFNg24h_vs_VCaP_NTC_IFNg24h <- read_tsv(paste(data_folder,"VCaPCHD1g2MAP3K7g2_IFNg24h_vs_VCaPNTC_IFNg24h.tsv", sep="/"), show_col_types = F)
#Comparison <- VCaP_MAP3K7g2_CTRL_vs_VCaP_NTC_CTRL <- read_tsv(paste(data_folder,"VCaPMAP3K7g2_CTRL_vs_VCaPNTC_CTRL.tsv", sep="/"), show_col_types = F)
#Comparison <- VCaP_MAP3K7g2_IFNg24h_vs_VCaP_NTC_IFNg24h <- read_tsv(paste(data_folder,"VCaPMAP3K7g2_IFNg24h_vs_VCaPNTC_IFNg24h.tsv", sep="/"), show_col_types = F)
#Comparison <- VCaP_MAP3K7g1_CTRL_vs_VCaP_NTC_CTRL <- read_tsv(paste(data_folder,"VCaPMAP3K7g1_CTRL_vs_VCaPNTC_CTRL.tsv", sep="/"), show_col_types = F)
#Comparison <- VCaP_MAP3K7g1_IFNg24h_vs_VCaP_NTC_IFNg24h <- read_tsv(paste(data_folder,"VCaPMAP3K7g1_IFNg24h_vs_VCaPNTC_IFNg24h.tsv", sep="/"), show_col_types = F)

Comparison$LogFoldChange <- Comparison$log2FoldChange * log(2)

Comparison$Tvalue <- Comparison$neg_log10_pval * Comparison$LogFoldChange

#load normalised counts
counts <- read_tsv(paste(data_folder, "VCaP_DKO_normalised_desq_counts.tsv", sep="/"), show_col_types = F)

#load metadata
samples <- read_csv(paste(data_folder, "VCaP_DKO_design.csv", sep="/"), col_types = NULL, show_col_types = FALSE) 
coldata <- samples[,c("replicate","condition","group","timepoint")]
rownames(coldata)<-coldata$replicate
sample <- samples$replicate
condition <- samples$condition

#combine to make list
design <- list(sample, condition)

data <- list(counts, design, Comparison)
names(data) <- c("counts", "design", "Deseq2")
```

#3. Extraction of normalised log transformed counts:
```{r}
# Remove NAs and set row names
counts <- data$counts %>%
  dplyr::mutate_if(~ any(is.na(.x)), ~ if_else(is.na(.x),0,.x)) %>% 
  column_to_rownames(var = "gene") %>% 
  as.matrix()
head(counts)
```


#4. Creating design metadata:
```{r}
design <- data$design
design
```


#5. Adding t values:
```{r}
# Extract t-values per gene
deg <- data$Deseq2 %>%
    select(gene, Tvalue) %>% 
    filter(!is.na(Tvalue)) %>% 
    column_to_rownames(var = "gene") %>%
    as.matrix()
head(deg)
```


#6. Using the PROGENy model and importing:
```{r}
net <- get_progeny(organism = 'human', top = 500)
net
```


#7. Conducting Multivarient Linear Model (MLM) analysis: 
```{r}
# Run mlm
sample_acts <- run_mlm(mat=counts, net=net, .source='source', .target='target', .mor='weight', minsize = 5)
sample_acts
```



#7. Visualisation:
```{r}
# Transform to wide matrix
sample_acts_mat <- sample_acts %>%
  pivot_wider(id_cols = 'condition', names_from = 'source',
              values_from = 'score') %>%
column_to_rownames('condition') %>%
  as.matrix()

# Scale per feature
sample_acts_mat <- scale(sample_acts_mat)

# Choose color palette
palette_length = 100
my_color = colorRampPalette(c("Darkblue", "white","red"))(palette_length)

my_breaks <- c(seq(-3, 0, length.out=ceiling(palette_length/2) + 1),
               seq(0.05, 3, length.out=floor(palette_length/2)))

# Plot
Pathway_heatmap_DKO_individuals <- pheatmap(sample_acts_mat, border_color = NA, color=my_color, breaks = my_breaks) 

#Save plot as eps
ggsave("Pathway_heatmap_VCaP_DKO_individuals.eps", plot = Pathway_heatmap_DKO_individuals$gtable, width = 8, height = 6, path = results_folder)
```

#8. Infer pathway activation:
```{r}
# Run mlm
contrast_acts <- run_mlm(mat=deg, net=net, .source='source', .target='target',
                  .mor='weight', minsize = 5)

#Compiling data for heatmap
#contrast_acts_1 <- contrast_acts$condition <- "CHD1g2_CTRL"
#contrast_acts_1 <- contrast_acts
#contrast_acts_2 <- contrast_acts$condition <- "CHD1g2_IFNg"
#contrast_acts_2 <- contrast_acts
#contrast_acts_3 <- contrast_acts$condition <- "CHD1g2MAP3K7g2_CTRL"
#contrast_acts_3 <- contrast_acts
#contrast_acts_4 <- contrast_acts$condition <- "CHD1g2MAP3K7g2_IFNg"
#contrast_acts_4 <- contrast_acts
#contrast_acts_5 <- contrast_acts$condition <- "MAP3K7g2_CTRL"
#contrast_acts_5 <- contrast_acts
#contrast_acts_6 <- contrast_acts$condition <- "MAP3K7g2_IFNg"
#contrast_acts_6 <- contrast_acts
#contrast_acts_7 <- contrast_acts$condition <- "MAP3K7g1_CTRL"
#contrast_acts_7 <- contrast_acts
#contrast_acts_8 <- contrast_acts$condition <- "MAP3K7g1_IFNg"
#contrast_acts_8 <- contrast_acts

#Group_comparison <- rbind(contrast_acts_1, contrast_acts_2, contrast_acts_3, contrast_acts_4, contrast_acts_7, contrast_acts_8)

#write.table(data.frame(Group_comparison), file=paste(results_folder,"VCaP_DKO_Pathway_scores_all_samples.tsv",sep="/"),sep="\t",row.names = FALSE)
```


#9. Plotting activity changes:
```{r}
# Plot
ggplot(contrast_acts, aes(x = reorder(source, score), y = score)) + 
    geom_bar(aes(fill = score), stat = "identity") +
    scale_fill_gradient2(low = "darkblue", high = "indianred", 
        mid = "whitesmoke", midpoint = 0) + 
    theme_minimal() +
    theme(axis.title = element_text(face = "bold", size = 12),
        axis.text.x = 
            element_text(angle = 45, hjust = 1, size =10, face= "bold"),
        axis.text.y = element_text(size =10, face= "bold"),
        panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank()) +
    xlab("Pathways")

Pathway_Plot_Name <- "VCaP_CHD1g2_CTRL.eps" 

ggsave(Pathway_Plot_Name, width = 12, height = 8.6, path = results_folder)
```


#10.Visualising the most active pathway - plotting of MAPK genes:
```{r}
#Type in pathways of interest to change plot
pathway <- 'JAK-STAT'

df <- net %>%
  filter(source == pathway) %>%
  arrange(target) %>%
  mutate(ID = target, color = "3") %>%
  column_to_rownames('target')
inter <- sort(intersect(rownames(deg),rownames(df)))
df <- df[inter, ]
df['t_value'] <- deg[inter, ]
df <- df %>%
  mutate(color = if_else(weight > 0 & t_value > 0, '1', color)) %>%
  mutate(color = if_else(weight > 0 & t_value < 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & t_value > 0, '2', color)) %>%
  mutate(color = if_else(weight < 0 & t_value < 0, '1', color))

#Plot
ggplot(df, aes(x = weight, y = t_value, color = color)) + geom_point() +
  scale_colour_manual(values = c("red","royalblue3","grey")) +
  geom_label_repel(aes(label = ID)) + 
  theme_minimal() +
  theme(legend.position = "none") +
  geom_vline(xintercept = 0, linetype = 'dotted') +
  geom_hline(yintercept = 0, linetype = 'dotted') +
  ggtitle(pathway)
```


#11. PLotting all sample groups against one another - heat map:
```{r}
#Load in pathways scores for all sample comparisons
Grouped_comparisons <- read_tsv(paste(data_folder, "VCaP_DKO_Pathway_scores_all_samples.tsv", sep = "/"), show_col_types = F)

# Transform to wide matrix
Grouped_comparisons_mat <- Grouped_comparisons %>%
  pivot_wider(id_cols = 'condition', names_from = 'source',
              values_from = 'score') %>%
  column_to_rownames('condition') %>%
  as.matrix()

# Choose color palette
palette_length = 100
my_color = colorRampPalette(c("Darkblue", "white","red"))(palette_length)

my_breaks <- c(seq(-15, 0, length.out=ceiling(palette_length/2) + 1),
               seq(0.05, 5, length.out=floor(palette_length/2)))

# Plot
Pathway_heatmap_DKO <- pheatmap(Grouped_comparisons_mat, border_color = NA, color=my_color, breaks = my_breaks) 
Pathway_heatmap_DKO

#Save plot as eps
ggsave("Pathway_heatmap_VCaP_DKO_grouped.eps", plot = Pathway_heatmap_DKO$gtable, width = 8, height = 6, path = results_folder)
```
